%-----------------------------------------------------------------------
% Beginning of chapter.tex
%-----------------------------------------------------------------------
%
%  This is a sample file for use with AMS-LaTeX.  It provides an example
%  of how to set up a file for a book to be typeset with AMS-LaTeX.
%
%  This is the driver file.  Separate chapters should be included at
%  the end of this file.
%
%  ***** DO NOT USE THIS FILE AS A STARTER FOR YOUR BOOK. *****
%  Follow the guidelines in the file chapter.template.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[11pt,a4paper]{memoir}
\chapterstyle{demo}
\epigraphfontsize{\small\itshape}
\setlength\epigraphwidth{8cm}
\setlength\epigraphrule{0pt}
\epigraphfontsize{\small\itshape}
%\includeonly{preface,chap1,biblio}

%\numberwithin{section}{chapter}
%\numberwithin{equation}{chapter}

% Include macros here
\input{macros}

%\renewcommand{\chaptername}{}
%\renewcommand{\thechapter}{}

\begin{document}
\numberwithin{section}{chapter}
\numberwithin{equation}{chapter}
\numberwithin{theorem}{chapter}

%\frontmatter
%\title{Convex Optimization}

%    Information for first author
%\author{Martin Lotz}
%    Address of record for the research reported here

%\address{School of Mathematics\\ The University of Manchester\\ Alan Turing Building, Oxford Road\\ Manchester M139PL}

%\email{martin.lotz@manchester.ed.uk}
%\urladdr{http://www.maths.manchester.ac.uk/\textasciitilde mlotz}

%\date{\today}

%\maketitle

%\newpage 

%\tableofcontents

%\include{preface}

\mainmatter

%\include{notation}

\section*{Overview}

\begin{enumerate}
\item Introduction (Lecture 1 + Introductory slides)
\begin{itemize}
 \item What is an optimization problem?
 \item Examples of optimization problems.
\end{itemize}
\item Unconstrained optimization and iterative algorithms (Lectures 2-6)
\begin{itemize}
 \item Unconstrained optimization
 \item Convex functions
 \item Gradient descent (steepest descent, step length, linear least squares)
 \item Convergence of iterative methods
 \item Convergence of gradient descent
 \item Newton's method
 \item Machine learning / supervised learning
\end{itemize}
\item Linear Programming Theory (Lectures 7-10)
\begin{itemize}
 \item Convex sets
 \item Linear programming duality: a first glance
 \item Polyhedra
 \item Farkas' Lemma
 \item Linear programming duality
 \item A first algorithm
 \item Optimality condition for linear programming
\end{itemize}

\end{enumerate}

\newpage

\section*{Questions and Answers}

\noindent {\em Q}: Which lectures are relevant for the test?\\

\noindent {\em A}: Weeks 1-5 (lectures 1-10). \\

\noindent {\em Q}: What do we have to know from the lectures?\\

\noindent {\em A}: The following {\em will not} be asked, though the statements should be known:
\begin{itemize}
\item Python code will not be part of the test;
\item Proofs in Section 4.3 on the convergence of gradient descent. What needs to be remembered from this section is that gradient descent has {\em linear} convergence (implied by the last Theorem in that section).
\item Proof of Theorem 5.6 (quadratic convergence of Newton's method), though the statement should be known.
\item Section 5.2 (Quasinewton methods)
\item Proof of Lemma 7.8.
\item Proof of Corollary 9.3.
\end{itemize}

As a rule of thumb, it is beneficial for the understanding to know the proof ideas, though precise reconstruction of proofs will not be asked.\\

\noindent {\em Q}: Which problems are relevant?\\

\noindent {\em A}: Problem that use Python or require cutting a shape out of paper will not be relevant to the test. More precisely, the following problems will not feature:

\begin{itemize}
\item Part B problems.
\item Problem 2.1(c) (the derivation).
\item Problem 2.3
\item Problem 3.2
\end{itemize}


%\include{lecture2}

%\appendix
%\include{appendix1}

%\backmatter
%\include{biblio}
%\include{index}
\end{document}

%-----------------------------------------------------------------------
% End of chapter.tex
%-----------------------------------------------------------------------
