\documentclass{article}
\include{macros}
\usepackage{pifont}

% Number of problem sheet
\newcounter{problemSheetNumber}
\setcounter{problemSheetNumber}{3}
\newcommand{\matlabprob}{\ding{100} \ }
\newcommand{\examprob}{\ding{80} \ }
%\setcounter{section}{\theproblemSheetNumber}  
%\renewcommand{\theparagraph}{(\thesection.\arabic{paragraph})}
\newcounter{problems}
\setcounter{problems}{0}
%\setlength{\parindent}{0cm}
\newcommand{\problem}[1]{\paragraph{(\theproblems)}\addtocounter{problems}{1}\label{#1}}
\ifx\marks\undefined
             \newcommand{\marks}[2][0mm]{\hspace{30mm}\mbox{}\vskip #1\hspace{-30mm}\hfill{\sf [#2 marks]}\\[-\baselineskip]}
\else
             \renewcommand{\marks}[2][0mm]{\hspace{30mm}\mbox{}\vskip #1\hspace{-30mm}\hfill{\sf [#2 marks]}\\[-\baselineskip]}
\fi
\newcommand{\solution}[1]{\paragraph{Solution (\theproblemSheetNumber.\theproblems)}\addtocounter{problems}{1}\label{#1}}

\pagestyle{fancy}
\lhead{MATH20602}
\chead{Numerical Analysis 1}
\rhead{March 17, 2014}

\begin{document} 
\begin{center}
{\Large {\bf Midterm test}}
\end{center}

\begin{center} 
\emph{Closed book.\quad Attempt all questions.\quad Calculators permitted. \quad 12:00-12:50}\\
{\em Please write your name and student identity number on the front page.}
\end{center}

%-- 
\problem{p0} Evaluate the expression
\begin{equation*}
  x^2-1
\end{equation*}
for $x=1.001$ by computing every step to four significant figures, and determine the relative error of the solution.
Rewrite the above equation in order to avoid cancellation errors. 

\marks[-6mm]{2}

\newpage
\mbox{}
\newpage

\problem{p2} Let $L_i(x)$, $0\leq i\leq n$, denote the Lagrange basis functions associated to interpolation nodes $x_0,\dots,x_n$, with $x_i=2i+1$ for $i=0,\dots,n$. Show that for all $x\in \R$,
\begin{equation*}
  L_1(x)+2L_2(x)+3L_3(x)+\dots +nL_n(x) = \frac{x-1}{2}.
\end{equation*}\marks[-6mm]{2}

\newpage
\mbox{}
\newpage
%-- 
\problem{p2} 
Compute the polynomial $p(x)$ that  interpolates the data 
\begin{equation*} 
(x_i, y_i)= (0, 0), (1, 1), (2,4), (3,9) 
\end{equation*} 
by use of a divided difference table. What is the degree of the resulting polynomial?

In general, given $n+1$ interpolation points $(x_i,y_i)$, $0\leq i\leq n$, how many arithmetic operations are needed to a) create the the divided difference table, and b) evaluate
the corresponding Newton interpolation polynomial?
\marks{3} 

\newpage
\mbox{}
\newpage

\problem{p3} Consider the function $f(x)=\cos(x)$ on the interval $[0,\pi]$. 
State the formula for the composite trapezium rule for approximating the integral $\int_{0}^{\pi} \cos(x) \ dx$.

Find, with justification, a number of intervals $n$ such that the integration error of the trapezium rule will be less than $10^{-2}$.
\marks{3} 

\newpage
\mbox{}
\newpage
 
\problem{p4} Define the degree of precision of a quadrature rule 
$I(f)$ that approximates $\int_a^b f(x)\,dx$. Determine a value of 
$\alpha$ such that the degree of precision of the following quadrature rule is at least $1$.
\begin{equation*} 
I(f)= \frac 13 \left( f(0)+ f(\alpha) +f(1)\right)\approx \int_0^1  f(x)\,dx.  
\end{equation*}
What is the degree of precision of the resulting quadrature rule? \marks{3} 

\newpage
\mbox{}
\newpage

\problem{p5}
Carry out two iterations of the Jacobi method for the system of equations
\begin{equation*}
  \begin{pmatrix} 2 & -1\\ -1 & 2 \end{pmatrix} \begin{pmatrix} x_1\\ x_2 \end{pmatrix} = \begin{pmatrix} 3\\3\end{pmatrix}
\end{equation*}
with starting vector $\vct{x}^{(0)} = (0,0)^{\top}$ (that is, $x_1^{(0)}=0, x_2^{(0)}=0$).

Compare the possible benefits and drawbacks of using the Gauss-Seidel method as opposed to the Jacobi method for large systems of equations.\marks{3}

\newpage
\mbox{}
\newpage

\problem{p6}
Define the operator norm of a matrix and show that a $2\times 2$ matrix $\mtx{A}$ satisfies the inequality
\begin{equation*}
\frac{1}{\sqrt{2}}\norm{\mtx{A}}_\infty \leq \norm{\mtx{A}}_2.
\end{equation*} 
You may use the fact that for $n$-vectors we have $\norm{\vct{x}}_\infty\leq \norm{\vct{x}}_2\leq \sqrt{n}\norm{\vct{x}}_\infty$. \marks{2}

\newpage
\mbox{}
\newpage

\problem{p7}
Determine the vector norms $\norm{\vct{x}}_2$ and $\norm{\vct{x}^{\top}}_1$ and the matrix norms $\norm{\mtx{A}}_\infty$ and $\norm{\mtx{A}}_1$ for
\begin{equation*}
  \vct{x} = \begin{pmatrix} -3\\ 2\\ 4 \end{pmatrix}, \quad
\mtx{A} =  \begin{pmatrix}
  1 & 0 & 1\\
  -2 & 4 & 3\\
  3 & -1 & 8
\end{pmatrix}.
\end{equation*}
Show that for any $n\times n$ matrix $\mtx{A}$, 
\begin{equation*}
  \norm{\mtx{A}}_1 = \norm{\mtx{A}^{\top}}_\infty,
\end{equation*}
where $\mtx{A}^{\top}$ denotes the matrix transpose.
\marks{3}

\newpage
\mbox{}
\newpage

\end{document} 
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
