\documentclass{article}
% Include macros here
\input{macros}
\usepackage{fancyhdr}
%\include{macros}
\usepackage{pifont}

% Number of problem sheet
\newcounter{problemSheetNumber}
\setcounter{problemSheetNumber}{1}
\newcommand{\matlabprob}{\ding{100} \ }
\newcommand{\examprob}{\ding{80} \ }
%\setcounter{section}{\theproblemSheetNumber}  
%\renewcommand{\theparagraph}{(\thesection.\arabic{paragraph})}
\newcounter{problems}
\setcounter{problems}{0}
%\setlength{\parindent}{0cm}
\renewcommand{\problem}{\paragraph{(\theproblemSheetNumber.\theproblems)}\addtocounter{problems}{1}}

%\theoremstyle{remark}
%\newtheorem{problem}[problemSheetNumber]{}

\pagestyle{fancy}
\lhead{MATH36061}
\chead{Convex Analysis}
\rhead{October 1, 2016}

\begin{document} 
\begin{center}
{\Large {\bf Problem Sheet \theproblemSheetNumber}}
\end{center}

Problems in Part A will be discussed in class. Problems in Part B come with solutions and should be tried at home, they will be discussed in class if time permits. 

\section*{Part A}

 \problem Find examples of
\begin{itemize}
 \item[(a)] A function $f\in C^2(\R)$ with a strict minimizer $x$ such that $f''(x)=0$ (that is, the second derivative is not positive definite).
 \item[(b)] A function $f\colon \R\to \R$ with a strict minimizer $x^*$ that is not an isolated local minimizer.
\end{itemize}

\problem For this problem you might want to recall some linear algebra.
\begin{itemize} 
 \item[(a)] Let $\mtx{A}\in \R^{n\times n}$ by a symmetric matrix, $\vct{b}\in \R^n$ and $c\in \R$. Show that the quadratic function
 \begin{equation}\label{eq:quad}\tag{1}
f(\vct{x}) = \frac{1}{2}\vct{x}^{\trans}\mtx{A}\vct{x}+\vct{b}^{\trans}\vct{x}+c  
 \end{equation}
 with symmetric $\mtx{A}$
is convex if and only if $\mtx{A}$ is positive semidefinite.
\item[(b)] Now let $\mtx{A}\in \R^{m\times n}$ by an arbitrary matrix. Show that the function
\begin{equation*}
 f(\vct{x}) = \norm{\mtx{A}\vct{x}-\vct{b}}_2^2
\end{equation*}
is convex (the $2$-norm is defined as $\norm{\vct{x}}_2=\vct{x}^{\trans}\vct{x}$). Moreover, if $m\geq n$ and the matrix $\mtx{A}$ has rank $m$, then it is strictly convex and the unique minimizer is
\begin{equation*}
 \vct{x}^* = (\mtx{A}^{\trans}\mtx{A})^{-1}\mtx{A}^{\trans}\vct{b}.
\end{equation*}
\end{itemize}

\problem A set $S\subseteq \R^n$ is called {\em convex}, if for any $\vct{x},\vct{y}\in S$ and $\lambda\in [0,1]$, 
\begin{equation*}
 \lambda \vct{x}+(1-\lambda)\vct{y}\in S.
\end{equation*}
In words, for any two points in $S$, the line segment joining them is also in $S$. Which of the following sets are convex?
\begin{itemize}
 \item[(a)] $S = \{\vct{x}\in \R^3 \mid \norm{\vct{x}}_2=1\}$ (the unit sphere);
 \item[(b)] $S = \{\vct{x}\in \R^2 \mid 1\leq x_1-x_2< 2\}$;
 \item[(c)] $S = \{\vct{x}\in \R^n \mid |x_1|+\cdots+|x_d|\leq 1\}$;
 \item[(d)] $S = \mathcal{S}^n_{+}\subset \R^{n\times n}$, the set of symmetric, positive semidefinite matrices.
\end{itemize}

\section*{Part B}

\problem 
In engineering applications\footnote{For example in the synthesis of linear time-invariant dynamical systems.} one sometimes encounters a problem of the form
\begin{equation}\label{eq:linfty}\tag{2}
 \minimize \norm{\mtx{A}\vct{x}-\vct{b}}_\infty,
\end{equation}
with $\mtx{A}\in \R^{m\times n}$, $\vct{b}\in \R^m$ and $\norm{\vct{x}}_\infty = \max_{1\leq i\leq n} |x_i|$ is the $\infty$-norm. 
\begin{itemize}
 \item[(a)] Draw the ``unit circle'' $\{\vct{x}\in \R^2 \mid \norm{\vct{x}}_\infty \leq 1\}$.
 \item[(b)] Formulate a linear programming problem $\mathcal{P}$ with decision variables $(\vct{x},t)$, such that if $(\vct{x}^*,t^*)$ is the unique minimizer of $\mathcal{P}$, then $\vct{x}^*$ is the unique minimizer of~\eqref{eq:linfty}. 
\end{itemize}
Even though~\eqref{eq:linfty} is not a linear programming problem (the objective is not linear), it is {\em equivalent} to one, in the sense that a minimizer can be read off the solution of a linear programming problem.

\problem For this problem we generalize the notion of convexity to function not necessarily defined on all of $\R^n$. Denote by $\dom f$ the {\em domain} of $f$, i.e., the set of $\vct{x}$ on which $f(\vct{x})$ attains a finite value. A function $f$ is called {\em convex}, if $\dom f$ is a convex set and for all $\vct{x},\vct{y}\in \dom f$ and $\lambda\in [0,1]$,
\begin{equation*}
 f(\lambda \vct{x}+(1-\lambda)\vct{y})\leq \lambda f(\vct{x})+(1-\lambda)f(\vct{y}).
\end{equation*}
Which of the following functions are convex?
\begin{itemize}
 \item[(a)] $f(x) = \log(x)$ on $\R_{++}$ (the positive real numbers);
 \item[(b)] $f(x) = x^4$ on $\R$;
 \item[(c)] $f(\vct{x}) = x_1x_2$ on $\R^2_{++}$;
 \item[(d)] $f(\vct{x}) = x_1/x_2$ on $\R^2_{++}$;
 \item[(e)] $f(x)=e^x-1$ on $\R$;
 \item[(f)] $f(\vct{x}) = \max_i x_i$ on $\R^n$.
\end{itemize}

\problem Using Python or another computing system, compute and plot the sequence of points $\vct{x}_k$, starting with $\vct{x}_0=(0,0)^{\trans}$, for the gradient descent algorithm for the problem
\begin{equation*}
 \minimize \norm{\mtx{A}\vct{x}-\vct{b}}_2^2
\end{equation*}
with data
 \begin{equation*}
  \mtx{A} = \begin{pmatrix}
             1 & 2\\
             2 & 1\\
             -1 & 0
            \end{pmatrix},
\quad \vct{b} = \begin{pmatrix}
                 10\\-1\\0
                \end{pmatrix}.
 \end{equation*}

\problem 

\end{document}
