\documentclass{article}
% Include macros here
\input{macros}
\usepackage{fancyhdr}
%\include{macros}
\usepackage{pifont}

% Number of problem sheet
\newcounter{problemSheetNumber}
\setcounter{problemSheetNumber}{3}
\newcommand{\matlabprob}{\ding{100} \ }
\newcommand{\examprob}{\ding{80} \ }
%\setcounter{section}{\theproblemSheetNumber}  
%\renewcommand{\theparagraph}{(\thesection.\arabic{paragraph})}
\newcounter{problems}
\setcounter{problems}{0}
%\setlength{\parindent}{0cm}
\renewcommand{\problem}{\paragraph{(\theproblemSheetNumber.\theproblems)}\addtocounter{problems}{1}}

%\theoremstyle{remark}
%\newtheorem{problem}[problemSheetNumber]{}

\pagestyle{fancy}
\lhead{MATH36061}
\chead{Convex Optimization}
\rhead{October 16, 2016}

\begin{document} 
\begin{center}
{\Large {\bf Problem Sheet \theproblemSheetNumber}}
\end{center}

Problems in Part A will be discussed in class. Problems in Part B come with solutions and should be tried at home. 

\section*{Part A}

\problem Recall that a norm was a function $\norm{\cdot}$ on $\R^n$ such that
\begin{itemize}
\item $\displaystyle \norm{\vct{x}}\geq 0$ and $\norm{\vct{x}}=0 \Leftrightarrow \vct{x}=\zerovct$;
\item $\displaystyle \norm{\lambda \vct{x}} = |\lambda|\norm{\vct{x}}$ for $\lambda\in\R$;
\item $\displaystyle \norm{\vct{x}+\vct{y}} \leq \norm{\vct{x}}+\norm{\vct{y}}$.
\end{itemize}
The most prominent examples are
\begin{equation*}
 \norm{\vct{x}}_2 = \sqrt{\sum_{i=1}^n x_i^2}, \quad \norm{\vct{x}}_1 = \sum_{i=1}^n |x_i|, \quad \norm{\vct{x}}_\infty = \max_{1\leq i\leq n} |x_i|.
\end{equation*}
\begin{itemize}
 \item[(a)] Show that for any norm, the {\em norm balls}
 \begin{equation*}
  B(\vct{p},r) := \{\vct{x} \mid \norm{\vct{x}-\vct{p}}\leq r\}
 \end{equation*}
are convex sets. 
\item[(b)] To a convex set $C$ with $\zerovct\in \inter C$ we can associate the {\em polar} set
\begin{equation*}
 C^* := \{\vct{y}\in \R^n \mid \forall \vct{x}\in C, \ip{\vct{x}}{\vct{y}}\leq 1\}.
\end{equation*}
Show that the polar of a convex set is again convex.
\item[(c)] Describe the polar sets of the norm balls $B(\zerovct,1)$ for the $1$, $2$, and $\infty$ norms.
\end{itemize}

\problem Let $C,D\subseteq \R^n$ be disjoint, non-empty, closed, bounded convex sets. Show that there exists an affine hyperplane $H$ strictly separating $C$ and $D$ (i.e., $C\subset \inter H_-$ and $D\subset \inter H_+$). Hint: consider the set $C-D=\{\vct{x}-\vct{y}\mid \vct{x}\in C, \vct{y}\in D\}$ and use the separation theorem for a convex set and a point from the lecture. Give an example of closed convex sets $C$ and $D$ that cannot be strictly separated by a hyperplane.

\problem (Facility location.) Facility location problems concern the location of points (facilities or devices on a circuit) in a region, some of which are required to be connected by links (streets or wires). The objective is to locate some of the points that are not fixed in a way that minimizes the transport cost along the links. The distance can be measured with respect to the usual Euclidean norm or the $1$-norm.

\begin{itemize}
 \item[(a)] Given a rectangular grid, show that the length of a path from coordiante $\vct{x}$ to $\vct{y}$ is given by
 \begin{equation*}
  d(\vct{x},\vct{y}) := \norm{\vct{x}-\vct{y}}_1.
 \end{equation*}
 \begin{figure}[h!]
\centering
\begin{tikzpicture}[thick,scale=1.5]
\node (A1) at (0,0)  [label=225:{$\vct{x}$}] {};
\node (A2) at (3,2)  [label=225:{$\vct{y}$}] {};
\draw[color=black!40] (0,0)--(3,0)--(3,2)--(0,2)--(0,0);
\draw[color=black!40] (0,1)--(3,1);
\draw[color=black!40] (0,2)--(3,2);
\draw[color=black!40] (1,0)--(1,2);
\draw[color=black!40] (2,0)--(2,2);

\draw[thick] (0,0)--(0,1)--(1,1)--(2,1)--(2,2)--(3,2);
\filldraw[black] (0,0) circle (2pt);
\filldraw[black] (3,2) circle (2pt);
\end{tikzpicture}
\caption{Manhattan distance}\label{fig:triangle}
\end{figure}
This is an example of the so-called {\em Manhattan distance}.
\item[(b)] Suppose now that we are given $N$ fixed points $\vct{p}_1,\dots,\vct{p}_N$ in on a fine grid in $\R^2$ and would like to find
a point $\vct{x}$ that minimizes the total $1$-norm distance
\begin{equation*}
 \minimize \sum_{i=1}^N \norm{\vct{p}_i-\vct{x}}_1 = \sum_{i=1}^N |p_{i,1}-x_1|+|p_{i,2}-x_2|.
\end{equation*}
Assume in addition that the point $\vct{x}$ is constrained to be in a polyhedral region
\begin{equation*}
 P = \{\vct{x} \mid \mtx{A}\vct{x}\leq \vct{b}\}.
\end{equation*}
Formulate this optimal placement problem as a linear programming problem.

\begin{figure}[h!]
\centering
\begin{tikzpicture}[thick,scale=1.5]
\node (A1) at (0,0)  [label=225:{$\vct{p}_1$}] {};
\node (A2) at (3,2)  [label=180:{$\vct{p}_2$}] {};
\node (A3) at (1,2)  [label=225:{$\vct{p}_3$}] {};


\draw[color=black!40,fill=blue!5] (1,0)--(2,0)--(2,1)--(1,1)--(1,0);
\draw (0,0)--(1.7,0.7);
\draw (1,2)--(1.7,0.7);
\draw (3,2)--(1.7,0.7);

\filldraw[black] (0,0) circle (2pt);
\filldraw[black] (1,2) circle (2pt);
\filldraw[black] (3,2) circle (2pt);
\filldraw[blue] (1.7,0.7) circle (2pt);
\node (A4) at (1.7,0.7) [label=270:{$\vct{x}$}] {};
\node (P) at (1,0) [label=45:{$P$}] {};
\end{tikzpicture}
\end{figure}
\end{itemize}

\section*{Part B}
\problem A {\em polyhedron} $P$ is a convex set defined by a system of linear inequalities
\begin{equation*}
 P = \{\vct{x}\in \R^n \mid \mtx{A}\vct{x}\leq \vct{b}\},
\end{equation*}
where $\mtx{A}\in \R^{m\times n}$, $\vct{x}\in \R^d$ and $\vct{b}\in \R^m$. A vertex $\vct{z}$ of $P$ is a point of $P$ that can not
be expressed as a convex combination of distinct points in $P$. 
For $\vct{z}\in P$, let $\mtx{A}_{\vct{z}}$ denote the matrix consisting of those rows $\vct{a}_i^{\trans}$ of $\mtx{A}$ where $\vct{a}_i^{\trans}\vct{z}=b_i$. For example, in Figure~\ref{fig:triangle} the matrix corresponding to the vertex $\vct{z}=(1,0)^{\trans}$ is given by
\begin{figure}[h!]
\centering
\begin{minipage}{0.3\textwidth}
\centering
\begin{tikzpicture}[thick,scale=1.5]
\draw[fill=blue!5] (-0,0)--(0,1)--(1,0)--(0,0);
%\draw[dashed,color=black!30,->] (0,0)--(1,1);
%\draw[dashed,color=black!30,->] (0,0)--(0,-1);
%\draw[dashed,color=black!30,->] (0,0)--(-1,0);
\end{tikzpicture}
\end{minipage}%
\begin{minipage}{0.7\textwidth}
\begin{equation*}
 \mtx{A} = \begin{pmatrix}
            1 & 1\\
            -1 & 0\\
            0 & -1\\
           \end{pmatrix}, \quad
 \vct{b} = \begin{pmatrix}
            1 \\ 0 \\ 0
           \end{pmatrix}, \quad
           \mtx{A}_{\vct{z}} = \begin{pmatrix}
                      1 & 1\\
                      0 & -1
                     \end{pmatrix}.
\end{equation*}
\end{minipage}
\caption{A triangle}\label{fig:triangle}
\end{figure}

Show that the vertices of $P$ are precisely the points $\vct{z}$ for which $\mtx{A}_{\vct{z}}$ has rank $n$. In particular, every polyhedron has only finitely many vertices.

\problem (Text classification) Consider the vocabulary consisting of the words



\end{document}
