\documentclass{article}
% Include macros here
\input{macros}
\usepackage{fancyhdr}
%\include{macros}
\usepackage{pifont}

% Number of problem sheet
\newcounter{problemSheetNumber}
\setcounter{problemSheetNumber}{7}
\newcommand{\matlabprob}{\ding{100} \ }
\newcommand{\examprob}{\ding{80} \ }
%\setcounter{section}{\theproblemSheetNumber}  
%\renewcommand{\theparagraph}{(\thesection.\arabic{paragraph})}
\newcounter{problems}
\setcounter{problems}{3}
%\setlength{\parindent}{0cm}
\renewcommand{\problem}[1]{\paragraph{(\theproblemSheetNumber.\theproblems)}\addtocounter{problems}{1}\label{#1}}
\renewcommand{\solution}[1]{\paragraph{Solution (\theproblemSheetNumber.\theproblems)}\addtocounter{problems}{1}\label{#1}}

\setcounter{MaxMatrixCols}{20}

\pagestyle{fancy}
\lhead{MATH36061}
\chead{Convex Optimization}
\rhead{November 25, 2016}

\begin{document} 
\begin{center}
{\Large {\bf Solutions to Part B of Problem Sheet \theproblemSheetNumber}}
\end{center}

\solution{pr:1} The Lagrangian of this problem is
\begin{align*}
 \mathcal{L}(\vct{x},\vct{\lambda},\vct{\mu}) &= \ip{\vct{c}}{\vct{x}}+\vct{\lambda}^{\trans}(\mtx{A}\vct{x}-\vct{b}) + \sum_{i=1}^d \mu_i x_i(1-x_i)\\
 &= \vct{x}^{\trans}\mathrm{diag}(\vct{\mu})\vct{x}+(\vct{c}+\mtx{A}^{\trans}\vct{\lambda}-\vct{\mu})^{\trans}\vct{x}-\vct{b}^{\trans}\vct{\lambda}.
\end{align*}
We want to minimize this over $\vct{x}$. If $\vct{\mu}<0$, then this function is unbounded below. Otherwise, we compute the minimum by setting the gradient to zero,
\begin{equation*}
 2\mathrm{diag}(\vct{\mu})\vct{x}+\vct{c}+\mtx{A}^{\trans}\vct{\lambda}-\vct{\mu} = \zerovct,
\end{equation*}
and get the expression (after resolving the above for $\vct{x}$ and plugging into the Lagrangian)
\begin{equation*}
 g(\vct{\lambda},\vct{\mu}) = \begin{cases}
                               -\vct{b}^{\trans}\vct{\lambda}-\frac{1}{4}\sum_{i=1}^d (c_i+\vct{a}_i^{\trans}\vct{\lambda}-\mu_i)^2/\mu_i, & \vct{\mu}\geq \zerovct\\
                               -\infty & \text{ else.}
                              \end{cases}
\end{equation*}
The dual problem is
\begin{align*}
 \maximize & -\vct{b}^{\trans}\vct{\lambda}-\frac{1}{4}\sum_{i=1}^d (c_i+\vct{a}_i^{\trans}\vct{\lambda}-\mu_i)^2/\mu_i\\
 \subjto & \vct{\mu}\geq \zerovct, \ \vct{\lambda}\geq \zerovct.
\end{align*}
The problem can be simplified a bit further by noting that for each summand
\begin{equation*}
 -\frac{1}{4}(c_i+\vct{a}_i^{\trans}\vct{\lambda}-\mu_i)^2/\mu_i,
\end{equation*}
if $c_i+\vct{a}_i^{\trans}\vct{\lambda}\geq 0$ we can set $\mu_i=c_i+\vct{a}_i^{\trans}\vct{\lambda}$ and the summand disappears, and if $c_i+\vct{a}_i^{\trans}\vct{\lambda}<0$, then we can maximize the summand by computing the derivative, and get the value $c_i+\vct{a}_i^{\trans}\vct{\lambda}$. Summarising, we have
\begin{align*}
 \maximize & -\vct{b}^{\trans}\vct{\lambda}+\sum_{i=1}^d \min\{0,c_i+\mtx{a}_i^{\trans}\vct{\lambda}\}\\
 \subjto & \vct{\lambda}\geq \zerovct.
\end{align*}
Note that the objective function is nonsmooth, but is still concave. One can reverse the signs (and replace the min with a max in the process) to obtain a convex problem. The optimal value of this new problem will be a lower bound, and approximation, to the optimal value of the problem we started out with.

%\solution{pr:1} The largest eigenvalue of $\mtx{A}$ can be written as
%\begin{equation*}
% \lambda_{\mathrm{max}}(\mtx{A}) = \max_{\vct{v}\in \R^d} \frac{\vct{v}^{\trans}\mtx{A}\vct{v}}{\vct{v}^{\trans}\vct{v}}.
%\end{equation*}
%Therefore, $t\geq \lambda_{\mathrm{max}}(\mtx{A})$ is equivalent to the statement that for all $\vct{v}$,
%\begin{equation*}
% t\geq \frac{\vct{v}^{\trans}\mtx{A}\vct{v}}{\vct{v}^{\trans}\vct{v}} \ \Leftrightarrow \ t\vct{v}^{\trans}\vct{v}-\vct{v}^{\trans}\mtx{A}\vct{v}\geq 0.
%\end{equation*}
%We can write $\vct{v}^{\trans}\vct{v}=\vct{v}^{\trans}\mtx{I}\vct{v}$, with the itendity matrix $\mtx{I}$, so that the above is equivalent to
%\begin{equation*}
% t\mtx{I}-\mtx{A}\succeq \zerovct.
%\end{equation*}
%The following semidefinite programming problem gives the largest eigenvalue:
%\begin{align*}
% \minimize & t\\
% \subjto & t\mtx{I}-\mtx{A}\succeq \zerovct.
%\end{align*}

\solution{pr:2} The idea here is to {\em relax} the problem: we replace the individual constraints $x_i^2=1$ with the weaker constraint that the sum of the squares is $m$,
\begin{equation}\label{eq:1}\tag{1}
  \minimize \vct{x}^{\trans}\mtx{W}\vct{x} \subjto \sum_{i=1}^n x_i^2 = n.
\end{equation}
This constraint includes the previous one, but we have more options. Therofore, the minimizer of~\eqref{eq:1} is smaller or equal to minimizer of the original problem, and any lower bound on this minimizer will be a lower bound on the minimizer of the original problem.

The Lagrangian of~\eqref{eq:1} is 
\begin{equation*}
  \mathcal{L}(\vct{x},\mu) = \vct{x}^{\trans}\mtx{W}\vct{x} + \mu (\vct{x}^{\trans}\vct{x}-n),
\end{equation*}
and computing the gradient, we see that the minimizer satisfies
\begin{equation*}
  \mtx{W}\vct{x} = -\mu \vct{x}.
\end{equation*}
This is precisely the equation for eigenvalues and eigenvectors of $\mtx{W}$! 
If we multiply the first equation with $\vct{x}^{\trans}$ and replace the term $\vct{x}^{\trans}\mtx{W}\vct{x}$ in the Lagrangian with $-\mu\vct{x}^{\trans}\vct{x}$, then we get
\begin{equation*}
  g(\mu) = -\mu n.
\end{equation*}
The largest possible such $\mu$ is arrived at when $\lambda = -\mu$ is the smallest eigenvalue of $\mtx{W}$. Therefore, $g(\mu) \geq n \lambda_{\text{min}}(\mtx{W})$, and in particular the optimal value of~\eqref{eq:1} (and of the original problem) is bounded by this.
\end{document}
